---
title: "assignment 2"
author: "Shruti Chanda, Shubham Chaudhary, Aarjav Sanghvi"
date: "10/18/2021"
output: html_document
---

Importing libraries
```{r}
install.packages("ggplotgui")
install.packages("randomForest")
install.packages("xgboost")

Pack <- c("tidyverse", "lubridate", "magrittr", "ggplot2", "pROC", "dplyr", "ggplotgui", "broom", "rpart", "C50", "ranger", "randomForest", "caret", "lift", "ROCR", "xgboost", "glmnet")

lapply(Pack, library, character.only = TRUE)
```

importing the updated dataset
```{r}
df <- read_csv("df3.csv")
glimpse(df)
```

manipulating data
```{r}
df[sapply(df, is.character)] <- lapply(df[sapply(df, is.character)],as.factor)
sapply(df,class)
```

splitting data
```{r}
TRNPROP = 0.5  #proportion of examples in the training sample
nr<-nrow(df)
trnIndex<- sample(1:nr, size = round(TRNPROP * nr), replace=FALSE)
lcdfTrn <- df[trnIndex, ]
lcdfTst <- df[-trnIndex, ]
```

Question 1- XGBoost
```{r}
fdum<-dummyVars(~.,data=df %>% select(-loan_status)) #do not include loan_status for this
dxlcdf <- predict(fdum, df)

levels(df$loan_status)
dylcdf <- class2ind(df$loan_status, drop2nd = FALSE) 
#fplcdf <- dycldf [ , 2] 
colcdf <- dylcdf [ , 1]

dxlcdfTrn <- dxlcdf[trnIndex,]
colcdfTrn <- colcdf[trnIndex]
dxlcdfTst <- dxlcdf[-trnIndex,]
colcdfTst <- colcdf[-trnIndex]

dxTrn <- xgb.DMatrix( subset(dxlcdfTrn, select=-c(annRet, actualTerm, actualReturn, total_pymnt)), label=colcdfTrn)
dxTst <- xgb.DMatrix( subset( dxlcdfTst,select=-c(annRet, actualTerm, actualReturn, total_pymnt)), label=colcdfTst)

xgbWatchlist <- list(train = dxTrn, eval = dxTst)

xgbParam <- list (max_depth = 5, eta = 0.01, objective = "binary:logistic",eval_metric="error", eval_metric = "auc")

xgb_lsM1 <- xgb.train( xgbParam, dxTrn, nrounds = 500, xgbWatchlist, early_stopping_rounds = 10 )


xgb_lsM1$best_iteration
 
xpredTrg<-predict(xgb_lsM1, dxTrn) # best_iteration is used
head(xpredTrg)

xgb_lsM1
 
table(pred=as.numeric(xpredTrg>0.5), act=colcdfTrn)

xpredTst<-predict(xgb_lsM1, dxTst)
pred_xgb_lsM1=prediction(xpredTst, lcdfTst$loan_status, label.ordering = c("Fully Paid","Charged Off"))
aucPerf_xgb_lsM1=performance(pred_xgb_lsM1, "tpr", "fpr")
plot(aucPerf_xgb_lsM1)
abline(a=0, b= 1)

xgbParam <- list (max_depth = 3, eta = 0.1, objective = "binary:logistic", eval_metric="error", eval_metric = "auc")

xgb_lscv <- xgb.cv( xgbParam, dxTrn, nrounds = 500, nfold=5, early_stopping_rounds = 10 )

xgb_lscv$best_iteration

best_cvIter <- which.max(xgb_lscv$evaluation_log$test_auc_mean)
xgb_lsbest <- xgb.train( xgbParam, dxTrn, nrounds = xgb_lscv$best_iteration )

xgb.importance(model = xgb_lsbest) %>% view()
xgbParam <- list (max_depth = 4, objective = "binary:logistic", eval_metric="error", eval_metric = "auc")
xgb_lsM1 <- xgb.train( xgbParam, dxTrn, nrounds = 500, xgbWatchlist, early_stopping_rounds = 10, eta=0.01 )

xgbParam <- list (max_depth = 4, objective = "binary:logistic", eval_metric="auc", eval_metric = "error")
xgb_lsM1 <- xgb.train( xgbParam, dxTrn, nrounds = 500, xgbWatchlist, early_stopping_rounds = 10, eta=0.01 )

xgb_lsM1 <- xgb.train( xgbParam, dxTrn, nrounds = 500, xgbWatchlist, early_stopping_rounds = 10, eta=0.5 )

xgbParam1<-list(objective = "binary:logistic", eval_metric="auc", eval_metric = "error")

xgb_lsM1 <- xgb.train( xgbParam1, dxTrn, nrounds = 500, xgbWatchlist, 
early_stopping_rounds = 10, eta=0.1, max_depth=6 )

xgb_lsM1 <- xgb.train( xgbParam1, dxTrn, nrounds = 1000, xgbWatchlist, 
early_stopping_rounds = 10, eta=0.1, max_depth=6, lambda=0.05 )

xgb_lsM1 <- xgb.train( xgbParam1, dxTrn, nrounds = 1000, xgbWatchlist, early_stopping_rounds = 10, eta=0.1, max_depth=6, lambda=0.05, subsample=0.7, colsample_bytree=0.5 )

xgb_lsM1 <- xgb.train( xgbParam1, dxTrn, nrounds = 1000, xgbWatchlist, early_stopping_rounds
= 10, eta=0.01, max_depth=6, subsample=0.7, colsample_bytree=0.5 )

xgbParamGrid <- expand.grid(max_depth = c(2, 5), eta = c(0.001, 0.01, 0.1) )

```


Question 2- GLM model -> LASSO REGRESSION
```{r}
#LASSO REGRESSION
xD<-lcdfTrn %>% select(-loan_status, -actualTerm, -annRet, -actualReturn, -total_pymnt)
yD<- lcdfTrn$actualReturn

glmDefault_cv<- cv.glmnet(data.matrix(xD), lcdfTrn$loan_status, family="binomial")
predDefSc <- predict(glmDefault_cv, data.matrix(lcdfTst %>% select(-loan_status, -actualTerm, -annRet, -actualReturn, -total_pymnt)), s="lambda.1se", type="class")
glmDefault_cv$lambda.min
glmDefault_cv$lambda.1se
tidy(coef(glmDefault_cv, s = glmDefault_cv$lambda.1se))
plot(glmDefault_cv)
plot(glmDefault_cv$glmnet.fit, xvar="lambda")
plot(glmDefault_cv$glmnet.fit)
summary(glmDefault_cv)
```

Question 2- GLM model -> RIDGE REGRESSION
```{r}
#Ridge Regression
yTrn<-factor(if_else(lcdfTrn$loan_status=="Fully Paid", '1', '0') )
xDTrn<-lcdfTrn%>% select(-loan_status, -actualTerm, -annRet, -actualReturn, -total_pymnt, -issue_d)
glmls_cv_a0<- cv.glmnet(data.matrix(xDTrn), lcdfTrn$loan_status, family="binomial", alpha=0)
plot(glmls_cv_a0)
```

Question 2- GLM model -> ELASTIC-NET REGRESSION
```{r}
#Elastic-Net Regression
glmls_cv_a2<- cv.glmnet(data.matrix(xDTrn), lcdfTrn$loan_status, family="binomial", alpha=0.2)
plot(glmls_cv_a2)
glmls_a5<- glmnet(data.matrix(xDTrn), lcdfTrn$loan_status, family="binomial", alpha=0.5)
plot(glmls_a5)
```

Random Forest for comparison purposes in various questions
```{r}
rfModel_Ret <- ranger(actualReturn ~., data=subset(lcdfTrn, select=-c(annRet, actualTerm, loan_status)), num.trees =200, 
importance='permutation')
rfPredRet_trn<- predict(rfModel_Ret, lcdfTrn)
sqrt(mean( (rfPredRet_trn$predictions - lcdfTrn$actualReturn)^2))
#sqrt(mean( ( (predict(rfModel_Ret, lcdfTst))$predictions - lcdfTst$actualReturn)^2))
plot ( (predict(rfModel_Ret, lcdfTst))$predictions, lcdfTst$actualReturn)
plot ( (predict(rfModel_Ret, lcdfTrn))$predictions, lcdfTrn$actualReturn)
```
Question 3
```{r}
xpredTst<-predict(xgb_lsM1, dxTst)
scoreTst_xgb_ls <- lcdfTst %>% select(grade, loan_status, actualReturn, actualTerm, int_rate) %>% mutate(score=xpredTst)
scoreTst_xgb_ls <- scoreTst_xgb_ls %>% mutate(tile=ntile(-score, 10))
scoreTst_xgb_ls %>% group_by(tile) %>% summarise(count=n(), avgSc=mean(score), numDefaults=sum(loan_status=="Charged Off"), 
avgActRet=mean(actualReturn), minRet=min(actualReturn), maxRet=max(actualReturn), avgTer=mean(actualTerm), totA=sum(grade=="A"), 
totB=sum(grade=="B" ), totC=sum(grade=="C"), totD=sum(grade=="D"), totE=sum(grade=="E"), totF=sum(grade=="F") )
```

Question 3- performance by decile
```{r}
#Performance by deciles
predRet_Trn <- lcdfTrn %>% select(grade, loan_status, actualReturn, actualTerm, int_rate) %>% mutate(predRet=(predict(rfModel_Ret, lcdfTrn))$predictions)
predRet_Trn <- predRet_Trn %>% mutate(tile=ntile(-predRet, 10))
predRet_Trn %>% group_by(tile) %>% summarise(count=n(), avgpredRet=mean(predRet), numDefaults=sum(loan_status=="Charged Off"), 
avgActRet=mean(actualReturn), minRet=min(actualReturn), maxRet=max(actualReturn), avgTer=mean(actualTerm), totA=sum(grade=="A"), totB=sum(grade=="B" ), totC=sum(grade=="C"), totD=sum(grade=="D"), totE=sum(grade=="E"), totF=sum(grade=="F") )
```

Question 4- Loan status 
```{r}
#Predict loan status
xpredTst<-predict(xgb_lsM1, dxTst)
scoreTst_xgb_ls <- lcdfTst %>% select(grade, loan_status, actualReturn, actualTerm, int_rate) %>% mutate(score=xpredTst)
scoreTst_xgb_ls <- scoreTst_xgb_ls %>% mutate(tile=ntile(-score, 10))
scoreTst_xgb_ls %>% group_by(tile) %>% summarise(count=n(), avgSc=mean(score), numDefaults=sum(loan_status=="Charged Off"), 
avgActRet=mean(actualReturn), minRet=min(actualReturn), maxRet=max(actualReturn), avgTer=mean(actualTerm), totA=sum(grade=="A"), 
totB=sum(grade=="B" ), totC=sum(grade=="C"), totD=sum(grade=="D"), totE=sum(grade=="E"), totF=sum(grade=="F") )

#test for above
#colcdf1 <- subset(colcdfTst,select=-c(annRet, actualTerm, total_pymnt, actualReturn))
predXgbRet_Tst <- lcdfTst %>% select(grade, loan_status, actualReturn, actualTerm, int_rate) %>% mutate(predXgbRet=predict(xgb_lsM1, subset(colcdfTst,select=-c(annRet, actualTerm, total_pymnt, actualReturn))) )
predXgbRet_Tst <- predXgbRet_Tst %>% mutate(tile=ntile(-predXgbRet, 10))
predXgbRet_Tst %>% group_by(tile) %>% summarise(count=n(), avgPredRet=mean(predXgbRet), 
numDefaults=sum(loan_status=="Charged Off"), avgActRet=mean(actualReturn), minRet=min(actualReturn), 
maxRet=max(actualReturn), avgTer=mean(actualTerm), totA=sum(grade=="A"), totB=sum(grade=="B" ), 
totC=sum(grade=="C"), totD=sum(grade=="D"), totE=sum(grade=="E"), totF=sum(grade=="F") )

#Consider top d deciles from M2, ranked by M1 scores
pRetSc <- predXgbRet_Tst %>% mutate(poScore=scoreTst_xgb_ls$score)
pRet_d <- pRetSc %>% filter(tile<=d)
pRet_d<- pRet_d %>% mutate(tile2=ntile(-poScore, 20))
pRet_d %>% group_by(tile2) %>% summarise(count=n(), avgPredRet=mean(predXgbRet), 
numDefaults=sum(loan_status=="Charged Off"), avgActRet=mean(actualReturn), minRet=min(actualReturn), 
maxRet=max(actualReturn), avgTer=mean(actualTerm), totA=sum(grade=="A"), totB=sum(grade=="B" ),
totC=sum(grade=="C"), totD=sum(grade=="D"), totE=sum(grade=="E"), totF=sum(grade=="F") )

#considering top d decile from M2
pRet_d<- pRet_d %>% mutate(expRet=predXgbRet*poScore)
pRet_d<- pRet_d %>% mutate(tile2=ntile(-expRet, 20))
pRet_d %>% group_by(tile2) %>% summarise(count=n(), avgPredRet=mean(predXgbRet), 
numDefaults=sum(loan_status=="Charged Off"), avgActRet=mean(actualReturn), minRet=min(actualReturn), 
maxRet=max(actualReturn), avgTer=mean(actualTerm), totA=sum(grade=="A"), totB=sum(grade=="B" ), 
totC=sum(grade=="C"), totD=sum(grade=="D"), totE=sum(grade=="E"), totF=sum(grade=="F") )

#Modeling loan_statuson lower grade loans - rf (ranger)

lg_lcdfTst<-lcdfTst %>% filter(grade=='C'| grade=='D'| grade== 'E'| grade== 'F'| grade== 'G')
rf_M1_lg <- ranger(loan_status ~., data=subset(lg_lcdfTrn, select=-c(annRet, actualTerm, actualReturn)), num.trees =200, 
probability=TRUE, importance='permutation') 
lg_scoreTstRF <- lg_lcdfTst %>% select(grade, loan_status, actualReturn, actualTerm, int_rate) %>% 
mutate(score=(predict(rf_M1_lg,lg_lcdfTst))$predictions[,"Fully Paid"])
lg_scoreTstRF <- lg_scoreTstRF %>% mutate(tile=ntile(-score, 10))
lg_scoreTstRF %>% group_by(tile) %>% summarise(count=n(), avgSc=mean(score), numDefaults=sum(loan_status=="Charged Off"), avgActRet=mean(actualReturn), minRet=min(actualReturn), 
maxRet=max(actualReturn), avgTer=mean(actualTerm), totA=sum(grade=="A"), totB=sum(grade=="B" ), 
totC=sum(grade=="C"), totD=sum(grade=="D"), totE=sum(grade=="E"), totF=sum(grade=="F"))


```

Question 5- ntile for glmnet
```{r}
predRet_Trn <- predRet_Trn %>% mutate(tile=ntile(-predRet, 10))
predRet_Trn %>% group_by(tile) %>% summarise(count=n(), avgpredRet=mean(predRet), numDefaults=sum(loan_status=="Charged Off"), 
avgActRet=mean(actualReturn), minRet=min(actualReturn), maxRet=max(actualReturn), avgTer=mean(actualTerm), totA=sum(grade=="A"), 
totB=sum(grade=="B" ), totC=sum(grade=="C"), totD=sum(grade=="D"), totE=sum(grade=="E"), totF=sum(grade=="F") )

```


